---
title: "IPF Trial run analysis on slide 0020866"
output:
  github_document:
    toc: true
  html_notebook:
    toc: true
---

This is a slide that was run in May 2024. The slide has punch biopsies arrayed, with 7 IPF samples and 3 normal samples. 

# Environment
```{r, results=F}
library(Seurat)
library(tidyverse)
```

# Read in Xenium 
```{r}
ipfArray1.xen <- LoadXenium('~/vcalab-files/datasets/IPF_xenium_trials_2023/output-XETG00143__0020866__Region_1__20240516__205940/')
```

## Add metadata 
This slide is a tissue array so need to specify which cells are from which piece of tissue, disease, etc.
In xenium viewer, the selection tool draws a region of interest and allows this region to be given a name.
Each core was identified in the Xenium viewer as "selections" and exported to csv file containing the cell barcodes
Selection file starts with 2 comment lines, selection name and Area. Will pull selection name as the item to ID

### helper function
```{r}
addSelectionNamesToXen <- function(csvlist, xenobj, meta.name) {
  # Accepts a list of filepaths pointing to CSV files from xenium viewer
  # CSV file should be exported from selection cells and first column should be cell barcode called Cell.ID
  # Assumes that selection name defined in xenium viewer is the name for Seurat object
  # Returns xenium object with added metadata variable with whatever supplied for meta.name
  lookup <- c()
  for (csvfile in csvlist) {
    name <- read.table(csvfile, nrows=1, header=F, comment.char='')$V3
    barcodes <- read.table(csvfile, sep=',', header=T)$Cell.ID
    lookup[barcodes] <- name
  }
  
  assignments <- unname(lookup[rownames(xenobj[[]])])
  assignments <- replace_na(assignments,'unassigned')
  xenobj[[meta.name]] <- assignments
  return(xenobj)
}
```

### Add sample metadata
```{r}
regions.csv.paths <- paste0('~/vcalab-files/datasets/IPF_xenium_trials_2023/output-XETG00143__0020866__Region_1__20240516__205940/regions/',
                            list.files('~/vcalab-files/datasets/IPF_xenium_trials_2023/output-XETG00143__0020866__Region_1__20240516__205940/regions', pattern='.csv')
)

ipfArray1.xen <- addSelectionNamesToXen(regions.csv.paths, ipfArray1.xen, 'Sample')
```

### Add disease metadata
```{r}
ipfArray1.xen$Dx <- unlist(lapply(ipfArray1.xen$Sample, function(x){
  if(startsWith(x,'IPF')) {return('IPF')} else if(startsWith(x,'N')) {return('Control')} else {return('unassigned')}
}))
```



# Identify cells as if normal single cell

An important caveat to this dataset is that these cells all come from different tissues and may need some kind of batch correction. 

## Baseline characteristics by sample
```{r}
VlnPlot(ipfArray1.xen, features='nCount_Xenium', group.by = 'Sample', pt.size = 0, y.max=500)
VlnPlot(ipfArray1.xen, features='nFeature_Xenium', group.by = 'Sample', pt.size = 0, y.max=150)
VlnPlot(ipfArray1.xen, features='nCount_ControlProbe', group.by = 'Sample', pt.size = 0)
```


## Subset somewhat strictly on feature count
On the theory that if there was only few features, but they had counts, it is probably still interpretable.
```{r}
ipfArray1.filtered.xen <- subset(ipfArray1.xen, subset=nCount_Xenium>20 & nFeature_Xenium>4)
```

How many cells did we filter out?
```{r}
print('original dataset:')
print(ncol(ipfArray1.xen))
print('filtered dataset:')  
print(ncol(ipfArray1.filtered.xen))
```

### Check characteristics of removed cells
```{r}
losers.xen <- subset(ipfArray1.xen, subset=nCount_Xenium<=20 | nFeature_Xenium<=4)
```
```{r}
print('Number of cells removed by sample')
table(losers.xen$Sample)
print('Proportion of cells removed by sample')
table(losers.xen$Sample)/ncol(losers.xen)
print('Proportion of cells in original dataset')
table(ipfArray1.xen$Sample)/ncol(ipfArray1.xen)
```


## Analyze without batch/sample information
```{r}
ipfArray1.filtered.xen <- NormalizeData(ipfArray1.filtered.xen)
ipfArray1.filtered.xen <- FindVariableFeatures(ipfArray1.filtered.xen)
ipfArray1.filtered.xen <- ScaleData(ipfArray1.filtered.xen)
ipfArray1.filtered.xen <- RunPCA(ipfArray1.filtered.xen, features = rownames(ipfArray1.filtered.xen))
ipfArray1.filtered.xen <- RunUMAP(ipfArray1.filtered.xen, dims = 1:30)
```
```{r}
ElbowPlot(ipfArray1.filtered.xen)
DimHeatmap(ipfArray1.filtered.xen, dims = 1:3, cells = 100, balanced = TRUE)
DimHeatmap(ipfArray1.filtered.xen, dims = 4:6, cells = 100, balanced = TRUE)
DimHeatmap(ipfArray1.filtered.xen, dims = 7:9, cells = 100, balanced = TRUE)
```

```{r}
# modification to reduce cluster number. i think the sparsity of data means that pruned edges leads to more communities.
ipfArray1.filtered.xen <- FindNeighbors(ipfArray1.filtered.xen, reduction='pca', dims=1:30, prune.SNN=0) 
ipfArray1.filtered.xen <- FindClusters(ipfArray1.filtered.xen, resolution = 0.6)
```
```{r}
DimPlot(ipfArray1.filtered.xen, group.by = 'Sample')
DimPlot(ipfArray1.filtered.xen, group.by = 'Dx', order = 'Control')
DimPlot(ipfArray1.filtered.xen, label=T)
```


```{r, fig.width=6, fig.height=8}
FeaturePlot(ipfArray1.filtered.xen, features = c("nFeature_Xenium", "nCount_Xenium", "nCount_ControlProbe", "nFeature_ControlProbe"), min.cutoff='q1', max.cutoff='q90')
FeaturePlot(ipfArray1.filtered.xen, features = c("COL1A1", "CTHRC1", "INMT", "SCN7A", "TNFAIP3", "IFITM1"), min.cutoff='q1', max.cutoff='q90')
FeaturePlot(ipfArray1.filtered.xen, features = c("AGR2", "ITGB6", "XBP1", "ABCA3", "KRT8", "CLDN4"), min.cutoff='q1', max.cutoff='q90')
FeaturePlot(ipfArray1.filtered.xen, features = c("EPCAM", "IRF1", "VEGFA", "GDF15", "FKBP5", "PDGFRA"), min.cutoff='q1', max.cutoff='q90')
FeaturePlot(ipfArray1.filtered.xen, features = c("MKI67", "COL1A1", "TOP2A", "CDK1", "CENPF", "CCNB2"), min.cutoff='q1', max.cutoff='q90')
FeaturePlot(ipfArray1.filtered.xen, features = c("RGS5", "MYH11", "PI16", "CDK1", "POSTN", "CCNB2"), min.cutoff='q1', max.cutoff='q90')
FeaturePlot(ipfArray1.filtered.xen, features = c("MUC1", "ABCA3", "KRT5", "TP63", "SCGB3A2", "CCNB2"), min.cutoff='q1', max.cutoff='q90')
```

## Analyze with split layers (i.e., normalize within batch/sample)

```{r}
ipfArray1.filtered.xen[['Xenium']] <- split(ipfArray1.filtered.xen[['Xenium']], f=ipfArray1.filtered.xen$Sample)
```

Redo dimensional reduction etc.

```{r}
ipfArray1.filtered.xen <- NormalizeData(ipfArray1.filtered.xen)
ipfArray1.filtered.xen <- FindVariableFeatures(ipfArray1.filtered.xen)
ipfArray1.filtered.xen <- ScaleData(ipfArray1.filtered.xen)
ipfArray1.filtered.xen <- RunPCA(ipfArray1.filtered.xen, features = rownames(ipfArray1.filtered.xen))
ipfArray1.filtered.xen <- RunUMAP(ipfArray1.filtered.xen, dims = 1:30)
```

```{r}
# modification to reduce cluster number. i think the sparsity of data means that pruned edges leads to more communities.
ipfArray1.filtered.xen <- FindNeighbors(ipfArray1.filtered.xen, reduction='pca', dims=1:30, prune.SNN=0) 
ipfArray1.filtered.xen <- FindClusters(ipfArray1.filtered.xen, resolution = 0.6)
```
```{r}
DimPlot(ipfArray1.filtered.xen, group.by = 'Sample')
DimPlot(ipfArray1.filtered.xen, group.by = 'Dx', order = 'Control')
DimPlot(ipfArray1.filtered.xen, label=T)
```



## Batch adjustment by sample
```{r}
ipfArray1.filtered.xen <- IntegrateLayers(object = ipfArray1.filtered.xen, method = HarmonyIntegration, orig.reduction = "pca", new.reduction = "integrated.harmony",
                                          assay='Xenium',
                                          verbose = FALSE)
```
Will defer rejoining of layers for now.

```{r}
ipfArray1.filtered.xen <- RunUMAP(ipfArray1.filtered.xen, reduction='integrated.harmony', dims = 1:30)
```

```{r}
# modification to reduce cluster number. i think the sparsity of data means that pruned edges leads to more communities.
ipfArray1.filtered.xen <- FindNeighbors(ipfArray1.filtered.xen, reduction='integrated.harmony', dims=1:30, prune.SNN=0) 
ipfArray1.filtered.xen <- FindClusters(ipfArray1.filtered.xen, resolution = 0.6)
```

```{r}
DimPlot(ipfArray1.filtered.xen, group.by = 'Sample')
DimPlot(ipfArray1.filtered.xen, group.by = 'Dx', order = 'Control')
DimPlot(ipfArray1.filtered.xen, label=T)
```

```{r, fig.width=6, fig.height=8}
FeaturePlot(ipfArray1.filtered.xen, features = c("nFeature_Xenium", "nCount_Xenium", "nCount_ControlProbe", "nFeature_ControlProbe"), min.cutoff='q1', max.cutoff='q90')
FeaturePlot(ipfArray1.filtered.xen, features = c("COL1A1", "CTHRC1", "INMT", "SCN7A", "TNFAIP3", "IFITM1"), min.cutoff='q1', max.cutoff='q90')
FeaturePlot(ipfArray1.filtered.xen, features = c("AGR2", "ITGB6", "XBP1", "ABCA3", "KRT8", "CLDN4"), min.cutoff='q1', max.cutoff='q90')
FeaturePlot(ipfArray1.filtered.xen, features = c("EPCAM", "IRF1", "VEGFA", "GDF15", "FKBP5", "PDGFRA"), min.cutoff='q1', max.cutoff='q90')
FeaturePlot(ipfArray1.filtered.xen, features = c("MKI67", "COL1A1", "TOP2A", "CDK1", "CENPF", "CCNB2"), min.cutoff='q1', max.cutoff='q90')
FeaturePlot(ipfArray1.filtered.xen, features = c("RGS5", "MYH11", "PI16", "CDK1", "POSTN", "CCNB2"), min.cutoff='q1', max.cutoff='q90')
```


# Reference mapping by IPF Cell Atlas

## Read in IPF Cell Atlas GSE136831
```{r}
original.dir <- ('~/vcalab-files/datasets/GSE136831_IPF_Cell_Atlas/')
metadata <- read_tsv(paste0(original.dir,'GSE136831_AllCells.Samples.CellType.MetadataTable.txt.gz'))
metadata <- as.data.frame(metadata)
rownames(metadata) <- metadata$CellBarcode_Identity

features <- read_tsv(paste0(original.dir,'GSE136831_AllCells.GeneIDs.txt.gz'))
cells <- read_tsv(paste0(original.dir,'GSE136831_AllCells.cellBarcodes.txt.gz'), col_names = F)

counts <- Matrix::readMM(paste0(original.dir,'GSE136831_RawCounts_Sparse.mtx.gz'))
rownames(counts) <- features$HGNC_EnsemblAlt_GeneID
colnames(counts) <- cells$X1

ipfatlas.cells <- CreateSeuratObject(counts=counts, meta.data=metadata)

rm(counts,cells,features,metadata)
gc()
```
```{r}
# restrict genes in ipfatlas to those genes in xenium
ipfatlas.cells <- subset(ipfatlas.cells, features=rownames(ipfArray1.filtered.xen))
```


```{r}
# routine processing
ipfatlas.cells <- NormalizeData(ipfatlas.cells)
ipfatlas.cells <- FindVariableFeatures(ipfatlas.cells)
ipfatlas.cells <- ScaleData(ipfatlas.cells, features=rownames(ipfatlas.cells))
ipfatlas.cells <- RunPCA(ipfatlas.cells)
ipfatlas.cells <- RunUMAP(ipfatlas.cells, dims=1:30)
```


## Reference mapping

In previous iterations it was clear that the "Multiplet" group in the IPF cell atlas ended up corresponding to AT2/AT1 compartment in Xenium. So remove multiplet to avoid misassigning Xenium cells to artifacts in the atlas.

```{r}
ipfatlas.cells <- subset(ipfatlas.cells, subset=CellType_Category=='Multiplet', invert=T)
ipfatlas.cells <- subset(ipfatlas.cells, subset=Manuscript_Identity=='Multiplet', invert=T)
```


```{r}
ipfatlas.anchors <- FindTransferAnchors(reference = ipfatlas.cells, query = ipfArray1.filtered.xen, dims = 1:30,
                                        reference.reduction = "pca",
                                        features=rownames(ipfArray1.filtered.xen)) # restrict features used for prediction to genes in Xen dataset
# Add top level categories
predictions1 <- TransferData(anchorset = ipfatlas.anchors, refdata = ipfatlas.cells$CellType_Category, dims = 1:30)
ipfArray1.filtered.xen <- AddMetaData(ipfArray1.filtered.xen, metadata = predictions1$predicted.id, col.name='ipfatlas.CellType_Category')

# Add published labels
predictions2 <- TransferData(anchorset = ipfatlas.anchors, refdata = ipfatlas.cells$Manuscript_Identity, dims = 1:30)
ipfArray1.filtered.xen <- AddMetaData(ipfArray1.filtered.xen, metadata = predictions2$predicted.id, col.name='ipfatlas.Manuscript_Identity')

```

## Reference mapping results

```{r}
DimPlot(ipfArray1.filtered.xen, group.by = 'ipfatlas.CellType_Category')
ggplot(predictions1, aes(factor(predicted.id), prediction.score.max)) + geom_violin(aes(fill=factor(predicted.id)))
```
```{r}
DimPlot(ipfArray1.filtered.xen, group.by = 'ipfatlas.Manuscript_Identity')
ggplot(predictions2, aes(factor(predicted.id), prediction.score.max)) + geom_violin(aes(fill=factor(predicted.id)))
```

## Export for Xenium Viewer

```{r}
scratch <- tibble(cell_id = names(ipfArray1.filtered.xen$ipfatlas.CellType_Category),
                      group = ipfArray1.filtered.xen$ipfatlas.CellType_Category)
write.csv(scratch, 'ipfArray1.filtered.xen_ipfatlas.CellType_Category.csv')
```
```{r}
scratch <- tibble(cell_id = names(ipfArray1.filtered.xen$ipfatlas.Manuscript_Identity),
                      group = ipfArray1.filtered.xen$ipfatlas.Manuscript_Identity)
write.csv(scratch, 'ipfArray1.filtered.xen_ipfatlas.Manuscript_Identity.csv')
```


# Subset epithelium
```{r}
ipfArray1.epi.xen <- subset(ipfArray1.filtered.xen, idents=c(0,1,6,12,13,14))
```


# Subset fibroblasts
```{r}
ipfArray1.fib.xen <- subset(ipfArray1.filtered.xen, idents=c(0,1,6,12,13,14))
```
## Recluster and reintegrate

```{r}
ipfArray1.fib.xen <- NormalizeData(ipfArray1.fib.xen, verbose=F)
ipfArray1.fib.xen <- FindVariableFeatures(ipfArray1.fib.xen, verbose=F) # defaults to finding 2000, which would be all of them, but downstream expects this.
ipfArray1.fib.xen <- ScaleData(ipfArray1.fib.xen, verbose=F)
ipfArray1.fib.xen <- RunPCA(ipfArray1.fib.xen, npcs = 30, features = rownames(ipfArray1.fib.xen), verbose=F)
ipfArray1.fib.xen <- IntegrateLayers(object = ipfArray1.fib.xen, method = HarmonyIntegration, orig.reduction = "pca", new.reduction = "integrated.harmony",
                                          assay='Xenium',
                                          verbose = FALSE)
ipfArray1.fib.xen <- RunUMAP(ipfArray1.fib.xen, reduction='integrated.harmony', dims = 1:30)

# modification to reduce cluster number. i think the sparsity of data means that pruned edges leads to more communities.
ipfArray1.fib.xen <- FindNeighbors(ipfArray1.fib.xen, reduction='integrated.harmony', dims=1:30, prune.SNN=0) 
ipfArray1.fib.xen <- FindClusters(ipfArray1.fib.xen, resolution = 0.6)
```
```{r}
DimPlot(ipfArray1.fib.xen, group.by = 'Sample')
DimPlot(ipfArray1.fib.xen, group.by = 'Dx', order = 'Control')
DimPlot(ipfArray1.fib.xen, label=T)
```

```{r}
FeaturePlot(ipfArray1.fib.xen, features = c("COL1A1", "CTHRC1", "INMT", "SCN7A"), min.cutoff='q1', max.cutoff='q90', order=T)
FeaturePlot(ipfArray1.fib.xen, features = c("COL1A1", "VEGFD", "SAA1", "CYP4B1"), min.cutoff='q1', max.cutoff='q90', order=T)
FeaturePlot(ipfArray1.fib.xen, features = c("CXCL12", "SFRP2", "CXCL14", "TGFBI"), min.cutoff='q1', max.cutoff='q90', order=T)
FeaturePlot(ipfArray1.fib.xen, features = c('COL1A1', "MKI67", "TOP2A", "CDK1"), min.cutoff='q1', max.cutoff='q90', order=T)
FeaturePlot(ipfArray1.fib.xen, features = c('COL1A1', "CENPF", "CCNB2",'KIT'), min.cutoff='q1', max.cutoff='q90', order=T)
FeaturePlot(ipfArray1.fib.xen, features = c("RGS5", "MYH11", "PI16","POSTN"), min.cutoff='q1', max.cutoff='q90', order=T)
```

Drop clusters 6 and 8 and redo

```{r}
ipfArray1.fib.xen <- subset(ipfArray1.fib.xen, idents = c(6,8), invert=T)
```

```{r}
ipfArray1.fib.xen <- NormalizeData(ipfArray1.fib.xen, verbose=F)
ipfArray1.fib.xen <- FindVariableFeatures(ipfArray1.fib.xen, verbose=F) # defaults to finding 2000, which would be all of them, but downstream expects this.
ipfArray1.fib.xen <- ScaleData(ipfArray1.fib.xen, verbose=F)
ipfArray1.fib.xen <- RunPCA(ipfArray1.fib.xen, npcs = 30, features = rownames(ipfArray1.fib.xen), verbose=F)
ipfArray1.fib.xen <- IntegrateLayers(object = ipfArray1.fib.xen, method = HarmonyIntegration, orig.reduction = "pca", new.reduction = "integrated.harmony",
                                          assay='Xenium',
                                          verbose = FALSE)
ipfArray1.fib.xen <- RunUMAP(ipfArray1.fib.xen, reduction='integrated.harmony', dims = 1:30, verbose=F)

# modification to reduce cluster number. i think the sparsity of data means that pruned edges leads to more communities.
ipfArray1.fib.xen <- FindNeighbors(ipfArray1.fib.xen, reduction='integrated.harmony', dims=1:30, prune.SNN=0) 
ipfArray1.fib.xen <- FindClusters(ipfArray1.fib.xen, resolution = 1)
```

```{r}
DimPlot(ipfArray1.fib.xen, group.by = 'Sample', pt.size=1.5)
DimPlot(ipfArray1.fib.xen, group.by = 'Dx', order = 'Control')
DimPlot(ipfArray1.fib.xen, label=T)
```

```{r}
FeaturePlot(ipfArray1.fib.xen, features = c("COL1A1", "CTHRC1", "INMT", "SCN7A"), min.cutoff='q1', max.cutoff='q90', order=T, pt.size=2)
FeaturePlot(ipfArray1.fib.xen, features = c( "VEGFD","CYP4B1","ASPN","BCHE"), min.cutoff='q1', max.cutoff='q90', order=T, pt.size=2)
FeaturePlot(ipfArray1.fib.xen, features = c("CXCL12", "SFRP2", "CXCL14", "TGFBI"), min.cutoff='q1', max.cutoff='q90', order=T, pt.size=2)
FeaturePlot(ipfArray1.fib.xen, features = c('COL1A1', "MKI67", "TOP2A", "CDK1"), min.cutoff='q1', max.cutoff='q50', order=T, pt.size=2)
FeaturePlot(ipfArray1.fib.xen, features = c('COL1A1', "CENPF", "CCNB2",'KIT'), min.cutoff='q1', max.cutoff='q50', order=T, pt.size=2)
FeaturePlot(ipfArray1.fib.xen, features = c("RGS5", "MYH11", "PI16","POSTN"), min.cutoff='q1', max.cutoff='q90', order=T, pt.size=2)
```

### Name fibroblast clusters and export to Xenium Viewer

```{r}
ipfArray1.fib.xen <- RenameIdents(object=ipfArray1.fib.xen,
                                  `0`='Smooth Muscle',
                                  `1`='Fib:Alveolar1',
                                  `2`='Fib:Alveolar2',
                                  `3`='Fib:Mixed?',
                                  `4`='Fib:CTHRC1',
                                  `5`='Fib:Inflammatory',
                                  `6`='Pericyte1',
                                  `7`='Pericyte2',
                                  `8`='Pericyte3',
                                  `9`='Fib:Adventitial',
                                  `10`='Fib:Proliferating')
ipfArray1.fib.xen$named_clusters <- Idents(ipfArray1.fib.xen)
```

```{r}
scratch <- tibble(cell_id = names(ipfArray1.fib.xen$named_clusters),
                      group = ipfArray1.fib.xen$named_clusters)
write.csv(scratch, 'ipfArray1.fib.xen_named_clusters.csv')
```

### Create proliferation module score
```{r}
ipfArray1.fib.xen[["Xenium"]] <- JoinLayers(ipfArray1.fib.xen[["Xenium"]])
ipfArray1.fib.xen <- AddModuleScore(ipfArray1.fib.xen, features=c("CENPF", "CCNB2",'KIT',"MKI67", "TOP2A", "CDK1" ),
                                    assay='Xenium',name='Proliferation', ctrl=15)
```
```{r}
FeaturePlot(ipfArray1.fib.xen, features = c("Proliferation6"), min.cutoff='q1', max.cutoff='q70', order=T, pt.size=2)

```



```{r}
DimPlot(ipfArray1.fib.xen)
DimPlot(ipfArray1.fib.xen, group.by = 'Sample')
```
```{r, fig.width=6, fig.height=8}
FeaturePlot(ipfArray1.fib.xen, features=c('COL1A1','CTHRC1','SCN7A','PI16','RGS5', 'MYH11'),
            min.cutoff='q1', max.cutoff='q80', order=F)

FeaturePlot(ipfArray1.fib.xen, features=c('RGS5','MYH11','CNN1','SFRP2'),
            min.cutoff='q1', max.cutoff='q99', order=F)
FeaturePlot(ipfArray1.fib.xen, features=c('CCNA1','CCNB2','MKI67','CENPF'),
            min.cutoff='q1', max.cutoff='q99', order=F)
```


# Seurat built-in niche analysis

## Niche analysis based on mapped IPF Atlas identities 
```{r}
ipfArray1.filtered.xen <- BuildNicheAssay(object = ipfArray1.filtered.xen, fov = "fov", group.by = "ipfatlas.Manuscript_Identity",
    niches.k = 12, neighbors.k = 40)
gc()
```
```{r}
table(ipfArray1.filtered.xen$ipfatlas.Manuscript_Identity, ipfArray1.filtered.xen$niches)
```

```{r}
cropped.coords <- Crop(ipfArray1.xen[["fov"]], x = c(1000, 2000), y = c(8000, 9000), coords = "plot")
ipfArray1.filtered.xen[["fibfocus1"]] <- cropped.coords
# visualize cropped area with cell segmentations & selected molecules
DefaultBoundary(ipfArray1.filtered.xen[["fibfocus1"]]) <- "segmentation"
```

```{r}
ImageDimPlot(ipfArray1.filtered.xen, fov = "fibfocus1", axes = TRUE, border.color = "white", border.size = 0.1, cols = "polychrome",
    group.by='niches') + coord_flip() + scale_x_reverse()
```

## Meta-niches
As we've previously seen, Seurat niche discovery has a strong tendecy to find aggregates of cells of the same type, illustrated by the epithelial rim and the fibroblast core coming out as separate.
Maybe try a niche analysis on niches, to find maybe co-occurrence of aberrant basaloid niche with fibroblasts and whatever those immune clusters are.

```{r}
# stow cell niches in a safe place
ipfArray1.filtered.xen$cellniches <- ipfArray1.filtered.xen$niches
```
```{r}
ipfArray1.filtered.xen <- BuildNicheAssay(object = ipfArray1.filtered.xen, fov = "fov", group.by = "cellniches",
    niches.k = 8, neighbors.k = 200)
```
```{r}
ImageDimPlot(ipfArray1.filtered.xen, fov = "fibfocus1", axes = TRUE, border.color = "white", border.size = 0.1, cols = "polychrome",
    group.by='niches') + coord_flip() + scale_x_reverse()
```


# Examine data as if spatial

## Generate assignments file for Xenium Explorer
For version 1.2.0, the instructons say:

"Visualize custom cells groups in Xenium Explorer by importing a CSV file specifying cell ids and corresponding group names. Build a CSV file containing each cell id in the first column and the group the cell should be assigned to in the second column. A cell may only be assigned to one group. Not all cells in the dataset need to be included.

Ensure the columns have headers named “cell_id” and “group”."

This is basically the named_clusters list with a header and saved as a CSV.

```{r}
scratch <- tibble(cell_id = names(ipfArray1.xen$named_clusters),
                      group = ipfArray1.xen$named_clusters)
write.csv(scratch, 'Seurat_named_clusters.csv')
```

## Imaging within Seurat
Found a region within Xenium explorer and figured the coordinates.

```{r}
cropped.coords <- Crop(ipfArray1.xen[["fov"]], x = c(1400, 1900), y = c(6400, 7000), coords = "plot")
ipfArray1.xen[["zoom"]] <- cropped.coords
# visualize cropped area with cell segmentations & selected molecules
DefaultBoundary(ipfArray1.xen[["zoom"]]) <- "segmentation"
ImageDimPlot(ipfArray1.xen, fov = "zoom", axes = TRUE, border.color = "white", border.size = 0.1, cols = "polychrome",
    coord.fixed = FALSE, molecules = c("ITGB6", "GDF15", "CTHRC1", "ABCA3", "CLDN4"), nmols = 10000)
```

