---
title: "Celltype identification on 0020227"
output: 
  github_document:
    toc: true
  html_notebook:
    toc: true
---

# Environment
```{r, results=F}
library(Seurat)
library(tidyverse)
```

## Read in Xenium 
```{r, results=F}
ipf3.xen <- LoadXenium('output-XETG00143__0020227__Region_1__20231214__022306/')
```

# General processing on 0020227 

## Filter

On the theory that if there was only few features, but they had counts, it is probably still interpretable.
```{r, results=F}
ipf3.filtered.xen <- subset(ipf3.xen, subset=nCount_Xenium>20 & nFeature_Xenium>4)
```
```{r}
print('original dataset:')
print(ncol(ipf3.xen))
print('filtered dataset:')
print(ncol(ipf3.filtered.xen))
```
```{r}
# cleanup cleanup
rm(ipf3.xen)
gc()
```

## Standard Processing
```{r}
DefaultAssay(ipf3.filtered.xen) <- 'Xenium'
ipf3.filtered.xen <- NormalizeData(ipf3.filtered.xen, verbose=F)
ipf3.filtered.xen <- FindVariableFeatures(ipf3.filtered.xen, verbose=F) # defaults to finding 2000, which would be all of them, but downstream expects this.
ipf3.filtered.xen <- ScaleData(ipf3.filtered.xen, verbose=F)
ipf3.filtered.xen <- RunPCA(ipf3.filtered.xen, npcs = 30, features = rownames(ipf3.filtered.xen), verbose=F)
ipf3.filtered.xen <- RunUMAP(ipf3.filtered.xen, dims = 1:30, verbose=F)
```

## Modified Clustering
Modification to reduce cluster number. i think the sparsity of data means that pruned edges leads to more communities. Major change is the prune.SNN parameter.
```{r}
ipf3.filtered.xen <- FindNeighbors(ipf3.filtered.xen, reduction='pca', dims=1:30, prune.SNN=0, verbose=F) 
ipf3.filtered.xen <- FindClusters(ipf3.filtered.xen, resolution = 0.6, verbose=F)
```
```{r}
DimPlot(ipf3.filtered.xen, label=T)
```

# Proliferation (general)
10X core set includes: CCNA1, CCNB2, CDK1, CENPF, KIT, MKI67, PCNA, TOP2A.
```{r}
FeaturePlot(ipf3.filtered.xen, features=c('COL1A1','CTHRC1','SCN7A','PLA2G2A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('RGS5','MYH11','CNN1','SFRP2'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('CCNA1','CCNB2','CDK1','CENPF'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('KIT','MKI67','PCNA','TOP2A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
```
```{r}
p <- DotPlot(ipf3.filtered.xen, features=c('EPCAM','ITGB6',
                                           'COL1A1','CTHRC1','SCN7A','PLA2G2A','RGS5','MFAP5','PI16','MYH11','CNN1','SFRP2',
                                      'CCNA1','CCNB2','CDK1','CENPF','KIT','MKI67','PCNA','TOP2A'))
p + coord_flip() + scale_x_discrete(limits = rev)
```

# Fibroblast subset
```{r}
ipf3.fib.xen <- subset(ipf3.filtered.xen, idents=c(0,5,9,15))
DimPlot(ipf3.fib.xen, label=T)
```

As usual I think the cell cycle signal is strong enough that it forced all proliferative cells into cluster 15 across different cell types. Proceed and prune later.

## Recluster
```{r}
ipf3.fib.xen <- NormalizeData(ipf3.fib.xen, verbose=F)
ipf3.fib.xen <- FindVariableFeatures(ipf3.fib.xen, verbose=F) # defaults to finding 2000, which would be all of them, but downstream expects this.
ipf3.fib.xen <- ScaleData(ipf3.fib.xen, verbose=F)
ipf3.fib.xen <- RunPCA(ipf3.fib.xen, npcs = 30, features = rownames(ipf3.filtered.xen), verbose=F)
ipf3.fib.xen <- RunUMAP(ipf3.fib.xen, dims = 1:30, verbose=F)
```
```{r}
ipf3.fib.xen <- FindNeighbors(ipf3.fib.xen, reduction='pca', dims=1:30, prune.SNN=0) 
ipf3.fib.xen <- FindClusters(ipf3.fib.xen, resolution = 1)
```
```{r}
DimPlot(ipf3.fib.xen, label=T)
```
Those fucking proliferative cells, EPCAM+ and COL1+ lumped together in cluster 9. Really justifies why regression is useful. Try higher res first.
```{r}
ipf3.fib.xen <- FindClusters(ipf3.fib.xen, resolution = 2)
DimPlot(ipf3.fib.xen, label=T)
```


```{r}
FeaturePlot(ipf3.fib.xen, features=c('EPCAM','ITGB6','ABCA3','INMT'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('COL1A1','CTHRC1','SCN7A','PLA2G2A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('RGS5','MYH11','CNN1','SFRP2'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('CCNA1','CCNB2','CDK1','CENPF'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('KIT','MKI67','PCNA','TOP2A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
```
Whew. Drop cluster 19 and 7, these got roped in here probably because they are DATPs that express some collagen or proliferative epithelium. One is almost certainly AT2 (ABCA3/KRT8/ITGB6 positive) and the other is proliferative epithelial. 

```{r}
ipf3.fib.xen <- subset(x = ipf3.fib.xen, idents = c(7,19), invert = TRUE)
```

## Recluster after pruning
```{r}
ipf3.fib.xen <- NormalizeData(ipf3.fib.xen, verbose=F)
ipf3.fib.xen <- FindVariableFeatures(ipf3.fib.xen, verbose=F) # defaults to finding 2000, which would be all of them, but downstream expects this.
ipf3.fib.xen <- ScaleData(ipf3.fib.xen, verbose=F)
ipf3.fib.xen <- RunPCA(ipf3.fib.xen, npcs = 30, features = rownames(ipf3.filtered.xen), verbose=F)
ipf3.fib.xen <- RunUMAP(ipf3.fib.xen, dims = 1:30, verbose=F)
```
```{r}
ipf3.fib.xen <- FindNeighbors(ipf3.fib.xen, reduction='pca', dims=1:30, prune.SNN=0, verbose=F) 
ipf3.fib.xen <- FindClusters(ipf3.fib.xen, resolution = 1.5, verbose=F)
```
```{r}
DimPlot(ipf3.fib.xen, label=T)
```
```{r}
FeaturePlot(ipf3.fib.xen, features=c('COL1A1','CTHRC1','POSTN','SCN7A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('INMT','RGS5','MYH11','CNN1'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('SFRP2','PLA2G2A','MFAP5','PI16'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('LGR5','WNT5A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('CCNA1','CCNB2','CDK1','CENPF'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('KIT','MKI67','PCNA','TOP2A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('LTBP2','COL5A2','COL8A1','DPP6'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('APOD','MFAP5','FBN1','FGFR4'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('PCOLCE2','PDGFRA','SFRP2','SVEP1'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('WNT2','MEDAG','PIM1','RARRES1'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('THBS2'),
            min.cutoff='q25', max.cutoff='q90', order=T)
```
```{r, fig.width=7, fig.height=6}
p <- DotPlot(ipf3.fib.xen, features=unique(c('EPCAM','ITGB6',
                                      'COL1A1','CTHRC1','POSTN','COL8A1','SCN7A','INMT','RGS5','MYH11','CNN1','SFRP2',
                                      'PLA2G2A','MFAP5','PI16','LGR5','WNT5A','CCNA1','CCNB2','CDK1','CENPF',
                                      'KIT','MKI67','PCNA','TOP2A','LTBP2','COL5A2','COL8A1','DPP6',
                                      'APOD','MFAP5','FBN1','FGFR4','PCOLCE2','PDGFRA','SFRP2','SVEP1',
                                      'WNT2','MEDAG','PIM1','RARRES1','THBS2'))
)
p + coord_flip() + scale_x_discrete(limits = rev)
```
```{r}
ipf3.fib.xen <- RenameIdents(object = ipf3.fib.xen,
                             `0` = "SmoothMuscle",
                             `1` = "AlveolarFib",
                             `2` = "FibroticFib_CTHRC1low",
                             `3` = "Pericyte",
                             `4` = "LipoFib",
                             `5` = "SmoothMuscle",
                             `6` = "AlveolarFib",
                             `7` = "FibroticFib_CTHRC1high",
                             `9` = "ProliferatingFib",
                             `8` = "AdventitialFib",
                             `10` = "AlveolarFib",
                             `11` = "Pericyte",
                             `12` = "AlveolarFib",
                             `13` = "AlveolarFib"
                             )
ipf3.fib.xen$named_clusters <- Idents(ipf3.fib.xen)
DimPlot(ipf3.fib.xen, label=T)
```

## Marker learning
On the theory that there might be hidden markers of utility in the Xenium probeset that we could co-opt for next cycle.
```{r}
fib.markers <- FindAllMarkers(ipf3.fib.xen, only.pos=T)
```

```{r}
fib.markers %>% group_by(cluster) %>% slice_head(n=10) %>% ungroup() -> top10
top10
```
```{r, fig.width=8, fig.height=7}
# manually save this
p1 <- DoHeatmap(ipf3.fib.xen, features = top10$gene) + NoLegend()
ggsave('forChris-proliferation-on-0020227_files/figure-gfm/markers-heatmap.png', units='in', width=9, height=8)
```

## Generate assignments file for Xenium Explorer

For version 1.2.0, the instructons say:

“Visualize custom cells groups in Xenium Explorer by importing a CSV file specifying cell ids and corresponding group names. Build a CSV file containing each cell id in the first column and the group the cell should be assigned to in the second column. A cell may only be assigned to one group. Not all cells in the dataset need to be included.

Ensure the columns have headers named “cell_id” and “group”.”

This is basically the named_clusters list with a header and saved as a CSV.

```{r}
scratch <- tibble(cell_id = names(ipf3.fib.xen$named_clusters),
                      group = ipf3.fib.xen$named_clusters)
write.csv(scratch, 'Seurat_fib_named_clusters.csv')
```


# Epithelium

```{r}
DimPlot(ipf3.filtered.xen, label=T)
FeaturePlot(ipf3.filtered.xen, features=c('EPCAM','KRT7','KRT8','ABCA3'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('ITGB6','CLDN4','XBP1','GDF15'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('KRT5','ITGB4','TP63','KRT15'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('FOXJ1','CCDC78','BCAS1','TP73'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('CYP2F1','STEAP4','MUC5B','TSPAN8'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('PDPN','SCEL','AGER','SEMA3B'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('CDH1','ASCL3','CFTR','SEMA3C'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('PLA2G2A','WT1','HP','CFB'),
            min.cutoff='q25', max.cutoff='q90', order=T)
```
```{r}
ipf3.epi.xen <- subset(ipf3.filtered.xen, ident=c(1,4,6,13,15))
DimPlot(ipf3.epi.xen)
```

## Recluster

Initial goal will be to identify contaminating fibroblasts and drop them.

```{r}
ipf3.epi.xen <- NormalizeData(ipf3.epi.xen, verbose=F)
ipf3.epi.xen <- FindVariableFeatures(ipf3.epi.xen, verbose=F) # defaults to finding 2000, which would be all of them, but downstream expects this.
ipf3.epi.xen <- ScaleData(ipf3.epi.xen, verbose=F)
ipf3.epi.xen <- RunPCA(ipf3.epi.xen, npcs = 30, features = rownames(ipf3.filtered.xen), verbose=F)
ipf3.epi.xen <- RunUMAP(ipf3.epi.xen, dims = 1:30, verbose=F)
ipf3.epi.xen <- FindNeighbors(ipf3.epi.xen, reduction='pca', dims=1:30, prune.SNN=0, verbose=F) 
ipf3.epi.xen <- FindClusters(ipf3.epi.xen, resolution = 2, verbose=F) # increased resolution to clear fibroblasts
```

```{r}
DimPlot(ipf3.epi.xen, label=T)
FeaturePlot(ipf3.epi.xen, features=c('EPCAM','KRT7','KRT8','ABCA3'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('ITGB6','CLDN4','XBP1','GDF15'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('KRT17','ITGB4','TP63','KRT15'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('FOXJ1','CCDC78','BCAS1','TP73'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('CYP2F1','STEAP4','MUC5B','TSPAN8'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('PDPN','SCEL','AGER','SEMA3B'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('CDH1','ASCL3','CFTR','SEMA3C'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('PLA2G2A','WT1','HP','CFB'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('MKI67','CCNB2','TOP2A'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('INMT','COL1A1','CTHRC1','SCN7A'),
            min.cutoff='q1', max.cutoff='q99', order=T)
```
```{r}
p <- DotPlot(ipf3.epi.xen, features=c('COL1A1','SCN7A','INMT','EPCAM','ABCA3','KRT7','KRT8','ITGB6', 'MKI67','CCNB2','TOP2A'))
p + coord_flip() + scale_x_discrete(limits = rev)
```

## Prune fibs and recluster

Cluster 17 is probably the replicating fibroblasts, to be distinguished from cluster 14 which are replicating EPCAM+ COL1A1-low/neg. Also remove cluster 12 which I think is contaminated with SCN7A positive fibroblasts, this is sensible because of the whole segmentation issue. Meaning: we cannot be confident that these cells are actually epithelial or some hybrid alveolar epithelial / fibroblast due to bad segmentation.

```{r}
ipf3.epi.xen <- subset(ipf3.epi.xen, idents=c(17,12), invert=T)
```
```{r}
ipf3.epi.xen <- NormalizeData(ipf3.epi.xen, verbose=F)
ipf3.epi.xen <- FindVariableFeatures(ipf3.epi.xen, verbose=F) # defaults to finding 2000, which would be all of them, but downstream expects this.
ipf3.epi.xen <- ScaleData(ipf3.epi.xen, verbose=F)
ipf3.epi.xen <- RunPCA(ipf3.epi.xen, npcs = 30, features = rownames(ipf3.filtered.xen), verbose=F)
ipf3.epi.xen <- RunUMAP(ipf3.epi.xen, dims = 1:30, verbose=F)
ipf3.epi.xen <- FindNeighbors(ipf3.epi.xen, reduction='pca', dims=1:30, prune.SNN=0, verbose=F) 
ipf3.epi.xen <- FindClusters(ipf3.epi.xen, resolution = 1.2, verbose=F) # lower resolution to lump analysis
```
```{r}
DimPlot(ipf3.epi.xen, label=T)
FeaturePlot(ipf3.epi.xen, features=c('EPCAM','KRT7','KRT8','ABCA3'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('ITGB6','CLDN4','XBP1','GDF15'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('KRT17','ITGB4','TP63','KRT15'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('FOXJ1','CCDC78','BCAS1','TP73'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('CYP2F1','STEAP4','MUC5B','TSPAN8'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('PDPN','SCEL','AGER','SEMA3B'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('CDH1','ASCL3','CFTR','SEMA3C'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('PLA2G2A','WT1','HP','CFB'),
            min.cutoff='q1', max.cutoff='q99', order=T)
FeaturePlot(ipf3.epi.xen, features=c('MKI67','CCNB2','TOP2A'),
            min.cutoff='q1', max.cutoff='q99', order=T)
p <- DotPlot(ipf3.epi.xen, features=c('EPCAM',
                                      'ABCA3','ETV5','SFTPD','SFTA2','PLA2G4F',
                                      'PDPN','SCEL','SEMA3B','AGER','SLC15A2',
                                      'TP63','KRT15',
                                      'FOXJ1','CCDC78','BCAS1','TP73',
                                      'CYP2F1','STEAP4','MUC5B','TSPAN8','MUC5AC','ERN2','AGR2',
                                      'KRT7','KRT8','ITGB6','GDF15','CLDN4','LGALS3',
                                      'DDIT3','XBP1','ATF4',
                                      'MKI67','CCNB2','TOP2A'))
p + coord_flip() + scale_x_discrete(limits = rev)
```

## What is epi cluster 2?

This is still confusing because I don't know what cluster 2 is.
```{r}
epi.cluster2.markers <- FindMarkers(ipf3.epi.xen, ident.1 = 2)
```
```{r}
epi.cluster2.markers %>% arrange(-pct.1)
```

This is baffling, 56% of these cells express CD3... some of these are plasma/B cell markers too. This must be a lymphocytes in the region of epithelial cells that catch some of the epithelial markers due to segmentation blah blah blah. If it's preferentially in the scleroderma lung then it will make more sense. 
```{r}
ImageDimPlot(ipf3.epi.xen, cols = "polychrome", size = 0.75)
```
```{r}
ImageDimPlot(ipf3.epi.xen, cols = "polychrome", size = 0.75)
ImageFeaturePlot(ipf3.epi.xen, features = 'CD3D', axes = TRUE, max.cutoff = "q90")
ImageFeaturePlot(ipf3.epi.xen, features = 'CD4', axes = TRUE, max.cutoff = "q90")
ImageFeaturePlot(ipf3.epi.xen, features = 'MZB1', axes = TRUE, max.cutoff = "q90")
ImageFeaturePlot(ipf3.epi.xen, features = 'nFeature_Xenium', axes = TRUE, max.cutoff = "q90")
ImageFeaturePlot(ipf3.epi.xen, features = 'nCount_Xenium', axes = TRUE, max.cutoff = "q90")
```
So... present in scleroderma but really mostly on the left side of the IPF slice. Seems to correlate with low count regions. 
```{r}
VlnPlot(ipf3.epi.xen, c('nFeature_Xenium','nCount_Xenium'), pt.size=0)
```
Indeed cluster 2 has lower counts. But the same can be said of a lot of other clusters.

## Assign epithelial names
```{r}
ipf3.epi.xen <- RenameIdents(object = ipf3.epi.xen,
                             `0` = "Epi_FibroticDATP1",
                             `1` = "Epi_AT2",
                             `2` = "Epi_mixedlowcount",
                             `3` = "Epi_FibroticDATP2",
                             `4` = "Epi_AT1",
                             `5` = "Epi_secretory",
                             `6` = "Epi_ciliated",
                             `7` = "Epi_basal",
                             `8` = "Epi_transitional?",
                             `9` = "Epi_transitional?",
                             `10` = "Epi_proliferating",
                             `11` = "Epi_mesothelial"
                             )
ipf3.epi.xen$named_clusters <- Idents(ipf3.epi.xen)
DimPlot(ipf3.epi.xen, label=T)
```

