---
title: "Proliferating fibroblasts on 0020227 for Chris"
output: 
  github_document:
    toc: true
  html_notebook:
    toc: true
---
# Environment
```{r}
library(Seurat)
library(tidyverse)
```
## Read in Xenium 
```{r}
ipf3.xen <- LoadXenium('output-XETG00143__0020227__Region_1__20231214__022306/')
```
# General processing on 0020227 
## Filter
On the theory that if there was only few features, but they had counts, it is probably still interpretable.
```{r}
ipf3.filtered.xen <- subset(ipf3.xen, subset=nCount_Xenium>20 & nFeature_Xenium>4)
```
```{r}
print('original dataset:')
print(ncol(ipf3.xen))
print('filtered dataset:')
print(ncol(ipf3.filtered.xen))
```
```{r}
# cleanup cleanup
rm(ipf3.xen)
gc()
```

## Standard Processing
```{r}
DefaultAssay(ipf3.filtered.xen) <- 'Xenium'
ipf3.filtered.xen <- NormalizeData(ipf3.filtered.xen, verbose=F)
ipf3.filtered.xen <- FindVariableFeatures(ipf3.filtered.xen, verbose=F) # defaults to finding 2000, which would be all of them, but downstream expects this.
ipf3.filtered.xen <- ScaleData(ipf3.filtered.xen, verbose=F)
ipf3.filtered.xen <- RunPCA(ipf3.filtered.xen, npcs = 30, features = rownames(ipf3.filtered.xen), verbose=F)
ipf3.filtered.xen <- RunUMAP(ipf3.filtered.xen, dims = 1:30, verbose=F)
```
## Modified Clustering
Modification to reduce cluster number. i think the sparsity of data means that pruned edges leads to more communities. Major change is the prune.SNN parameter.
```{r}
ipf3.filtered.xen <- FindNeighbors(ipf3.filtered.xen, reduction='pca', dims=1:30, prune.SNN=0) 
ipf3.filtered.xen <- FindClusters(ipf3.filtered.xen, resolution = 0.6)
```
```{r}
DimPlot(ipf3.filtered.xen, label=T)
```
# Proliferation (general)
10X core set includes: CCNA1, CCNB2, CDK1, CENPF, KIT, MKI67, PCNA, TOP2A.
```{r}
FeaturePlot(ipf3.filtered.xen, features=c('COL1A1','CTHRC1','SCN7A','PLA2G2A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('RGS5','MYH11','CNN1','SFRP2'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('CCNA1','CCNB2','CDK1','CENPF'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.filtered.xen, features=c('KIT','MKI67','PCNA','TOP2A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
```
```{r}
p <- DotPlot(ipf3.filtered.xen, features=c('EPCAM','ITGB6',
                                           'COL1A1','CTHRC1','SCN7A','PLA2G2A','RGS5','MFAP5','PI16','MYH11','CNN1','SFRP2',
                                      'CCNA1','CCNB2','CDK1','CENPF','KIT','MKI67','PCNA','TOP2A'))
p + coord_flip() + scale_x_discrete(limits = rev)
```
```{r}
ImageFeaturePlot(ipf3.filtered.xen, features=c('ARG1'),
            min.cutoff='q25', max.cutoff='q60', cols=c('white','red'))
ImageFeaturePlot(ipf3.filtered.xen, features=c('IL13'),
            min.cutoff='q25', max.cutoff='q60', cols=c('white','red'))
ImageFeaturePlot(ipf3.filtered.xen, features=c('IL6'),
            min.cutoff='q25', max.cutoff='q60', cols=c('white','red'))
```
# Fibroblast subset
```{r}
ipf3.fib.xen <- subset(ipf3.filtered.xen, idents=c(0,5,9,15))
DimPlot(ipf3.fib.xen, label=T)
```
As usual I think the cell cycle signal is strong enough that it forced all proliferative cells into cluster 15 across different cell types. Proceed and prune later.
## Recluster
```{r}
ipf3.fib.xen <- NormalizeData(ipf3.fib.xen, verbose=F)
ipf3.fib.xen <- FindVariableFeatures(ipf3.fib.xen, verbose=F) # defaults to finding 2000, which would be all of them, but downstream expects this.
ipf3.fib.xen <- ScaleData(ipf3.fib.xen, verbose=F)
ipf3.fib.xen <- RunPCA(ipf3.fib.xen, npcs = 30, features = rownames(ipf3.filtered.xen), verbose=F)
ipf3.fib.xen <- RunUMAP(ipf3.fib.xen, dims = 1:30, verbose=F)
```
```{r}
ipf3.fib.xen <- FindNeighbors(ipf3.fib.xen, reduction='pca', dims=1:30, prune.SNN=0) 
ipf3.fib.xen <- FindClusters(ipf3.fib.xen, resolution = 1)
```
```{r}
DimPlot(ipf3.fib.xen, label=T)
```
Those fucking proliferative cells, EPCAM+ and COL1+ lumped together in cluster 9. Really justifies why regression is useful. Try higher res first.
```{r}
ipf3.fib.xen <- FindClusters(ipf3.fib.xen, resolution = 2)
DimPlot(ipf3.fib.xen, label=T)
```


```{r}
FeaturePlot(ipf3.fib.xen, features=c('EPCAM','ITGB6','ABCA3','INMT'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('COL1A1','CTHRC1','SCN7A','PLA2G2A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('RGS5','MYH11','CNN1','SFRP2'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('CCNA1','CCNB2','CDK1','CENPF'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('KIT','MKI67','PCNA','TOP2A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
```
Whew. Drop cluster 19 and 7, these got roped in here probably because they are DATPs that express some collagen or proliferative epithelium. One is almost certainly AT2 (ABCA3/KRT8/ITGB6 positive) and the other is proliferative epithelial. 

```{r}
ipf3.fib.xen <- subset(x = ipf3.fib.xen, idents = c(7,19), invert = TRUE)
```
## Recluster after pruning
```{r}
ipf3.fib.xen <- NormalizeData(ipf3.fib.xen, verbose=F)
ipf3.fib.xen <- FindVariableFeatures(ipf3.fib.xen, verbose=F) # defaults to finding 2000, which would be all of them, but downstream expects this.
ipf3.fib.xen <- ScaleData(ipf3.fib.xen, verbose=F)
ipf3.fib.xen <- RunPCA(ipf3.fib.xen, npcs = 30, features = rownames(ipf3.filtered.xen), verbose=F)
ipf3.fib.xen <- RunUMAP(ipf3.fib.xen, dims = 1:30, verbose=F)
```
```{r}
ipf3.fib.xen <- FindNeighbors(ipf3.fib.xen, reduction='pca', dims=1:30, prune.SNN=0, verbose=F) 
ipf3.fib.xen <- FindClusters(ipf3.fib.xen, resolution = 1.5, verbose=F)
```
```{r}
DimPlot(ipf3.fib.xen, label=T)
```
```{r}
FeaturePlot(ipf3.fib.xen, features=c('COL1A1','CTHRC1','POSTN','SCN7A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('INMT','RGS5','MYH11','CNN1'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('SFRP2','PLA2G2A','MFAP5','PI16'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('LGR5','WNT5A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('CCNA1','CCNB2','CDK1','CENPF'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('KIT','MKI67','PCNA','TOP2A'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('LTBP2','COL5A2','COL8A1','DPP6'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('APOD','MFAP5','FBN1','FGFR4'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('PCOLCE2','PDGFRA','SFRP2','SVEP1'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('WNT2','MEDAG','PIM1','RARRES1'),
            min.cutoff='q25', max.cutoff='q90', order=T)
FeaturePlot(ipf3.fib.xen, features=c('THBS2'),
            min.cutoff='q25', max.cutoff='q90', order=T)
```
```{r, fig.width=7, fig.height=6}
p <- DotPlot(ipf3.fib.xen, features=unique(c('EPCAM','ITGB6',
                                      'COL1A1','CTHRC1','POSTN','COL8A1','SCN7A','INMT','RGS5','MYH11','CNN1','SFRP2',
                                      'PLA2G2A','MFAP5','PI16','LGR5','WNT5A','CCNA1','CCNB2','CDK1','CENPF',
                                      'KIT','MKI67','PCNA','TOP2A','LTBP2','COL5A2','COL8A1','DPP6',
                                      'APOD','MFAP5','FBN1','FGFR4','PCOLCE2','PDGFRA','SFRP2','SVEP1',
                                      'WNT2','MEDAG','PIM1','RARRES1','THBS2'))
)
p + coord_flip() + scale_x_discrete(limits = rev)
```
```{r}
ipf3.fib.xen <- RenameIdents(object = ipf3.fib.xen,
                             `0` = "SmoothMuscle",
                             `1` = "AlveolarFib",
                             `2` = "FibroticFib_CTHRC1low",
                             `3` = "Pericyte",
                             `4` = "LipoFib",
                             `5` = "SmoothMuscle",
                             `6` = "AlveolarFib",
                             `7` = "FibroticFib_CTHRC1high",
                             `9` = "ProliferatingFib",
                             `8` = "AdventitialFib",
                             `10` = "AlveolarFib",
                             `11` = "Pericyte",
                             `12` = "AlveolarFib",
                             `13` = "AlveolarFib"
                             )
ipf3.fib.xen$named_clusters <- Idents(ipf3.fib.xen)
DimPlot(ipf3.fib.xen, label=T)
```
## Marker learning
On the theory that there might be hidden markers of utility in the Xenium probeset that we could co-opt for next cycle.
```{r}
fib.markers <- FindAllMarkers(ipf3.fib.xen, only.pos=T)
```

```{r}
fib.markers %>% group_by(cluster) %>% slice_head(n=10) %>% ungroup() -> top10
top10
```
```{r, fig.width=8, fig.height=7}
# manually save this
p1 <- DoHeatmap(ipf3.fib.xen, features = top10$gene) + NoLegend()
ggsave('forChris-proliferation-on-0020227_files/figure-gfm/markers-heatmap.png', units='in', width=9, height=8)
```
## Generate assignments file for Xenium Explorer

For version 1.2.0, the instructons say:

“Visualize custom cells groups in Xenium Explorer by importing a CSV file specifying cell ids and corresponding group names. Build a CSV file containing each cell id in the first column and the group the cell should be assigned to in the second column. A cell may only be assigned to one group. Not all cells in the dataset need to be included.

Ensure the columns have headers named “cell_id” and “group”.”

This is basically the named_clusters list with a header and saved as a CSV.

```{r}
scratch <- tibble(cell_id = names(ipf3.fib.xen$named_clusters),
                      group = ipf3.fib.xen$named_clusters)
write.csv(scratch, 'Seurat_fib_named_clusters.csv')
```
